---
title: "Differential expression analysis with diffwrap"
output: html
editor: visual
author: Vidal Fey (vidal.fey@tuni.fi)
date: 'Last updated: `r format(Sys.Date(), "%B %d, %Y")`'
format:
  html:
    code-fold: true
    toc: true
    html-math-method: katex
    toc-title: Contents
    number-sections: true
    number-depth: 3
    code-copy: true
vignette: >
  %\VignetteIndexEntry{Differential expression analysis with diffwrap}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| label: load-packages
#| include: false
 
library(diffwrap)
```

## Introduction

::: {style="text-align: justify;"}
The ***diffwrap*** package provides functionality for differential expression analysis of read counts from mRNA sequencing data or miRNA expression values generated by the CAP-miRSEQ *expression_reports.sh* script. The workflow follows the **edgeR**-**limma** expression data analysis pipeline providing options for different approaches, e.g., using only **edgeR** functions or tools available in the **limma** package such as `voom`. The functions in the package generate text files with differential expression lists, optionally annotated with information from *biomart*, expression summary plots as well as several QC plots.
:::

## Example data

::: {style="text-align: justify;"}
The data used in the code example were downloaded from ...

The study looked at ...
:::

## Analysis workflow

::: {style="text-align: justify;"}
In the simplest case
:::
