---
title: "Differential expression analysis with diffwrap"
output: html
editor: visual
author: Vidal Fey (vidal.fey@gmail.com)
date: 'Last updated: `r format(Sys.Date(), "%B %d, %Y")`'
format:
  html:
    code-fold: true
    toc: true
    html-math-method: katex
    toc-title: Contents
    number-sections: true
    number-depth: 3
    code-copy: true
vignette: >
  %\VignetteIndexEntry{Differential expression analysis with diffwrap}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: load-packages
#| include: false

library(diffwrap)
```

## Introduction

::: {style="text-align: justify;"}
The ***diffwrap*** package provides functionality for differential expression analysis of read counts from mRNA sequencing data or miRNA expression values generated by the CAP-miRSeq *expression_reports.sh* script. The workflow follows the **edgeR**-**limma** expression data analysis pipeline providing options for different approaches, e.g., using only **edgeR** functions or tools available in the **limma** package such as `voom`. The functions in the package generate text files with differential expression lists, optionally annotated with information from *biomart*, expression summary plots as well as several QC plots.

The main entry point is `diffExpr()`, a convenience wrapper that runs the whole pipeline with sensible defaults. Every step it performs is also available as an exported function, so the workflow can equally well be run manually, or individual steps reused in a different context. This vignette shows both routes.
:::

## Pipeline overview

::: {style="text-align: justify;"}
Pre-processing runs across the top of the diagram below (with the MDS overview as an optional side branch). A *Mode resolution* hub then fans into five colour-coded lanes, one per analysis mode: each lane shows the flags that select it (`pairs`, `block`, `do.voom`) and the resulting recipe (design · contrasts · fit). The lanes reconverge at contrast extraction, and the required pipeline ends at the DE tables; QC, per-contrast plots, Venn and enrichment hang off as optional outputs. The five modes are described in @sec-modes; enrichment (@sec-enrichment) is detailed separately in @fig-enrichment.
:::

![The diffwrap `diffExpr()` workflow. Pre-processing feeds a Mode resolution hub that fans into the five analysis modes (role × engine), each labelled with its selecting flags and its design/contrasts/fit recipe. The modes reconverge at contrast extraction; the required pipeline ends at the DE tables, with QC, plots, Venn and enrichment as optional branches.](diffwrap-pipeline.svg){#fig-pipeline}

## Example data

::: {style="text-align: justify;"}
The package ships a small **simulated** data set so that the workflow can be demonstrated without any external data or network access. It is *not* a real experiment: counts were drawn from a negative binomial model with a known ground truth, which makes it convenient for illustration and testing.

The design is eight samples in two groups of four, `control` and `treated`. Each of the four subjects (`P1`--`P4`) contributed one control and one treated sample, so the same data can be analysed as an unpaired comparison, as a paired design, or as a blocked design. A subject-specific offset is built into the simulation, giving the paired analyses a real effect to remove.

Of the 400 genes, 60 are truly differentially expressed (30 up, 30 down) with absolute log2 fold changes between 1.2 and 3. A further 40 genes are given deliberately low expression so that they straddle the default filtering threshold, and the five `htseq-count` summary rows (`__no_feature` and friends) are retained so that the filtering step has something to remove.
:::

::: {style="text-align: justify;"}
The package ships this data set as the lazy-loaded objects `diffwrap_counts` and `diffwrap_samp_info` (see `?diffwrap_counts`), and as the plain tab-separated files `example_counts.tsv` and `example_samp_info.tsv` under the package's `extdata` directory. So that this vignette is completely self-contained and reproducible, we regenerate an equivalent data set inline from the same simulation model rather than loading the shipped copy:
:::

```{r}
#| label: example-data
set.seed(20240721)

n_genes <- 400L; n_true_de <- 60L
samples <- sprintf("S%02d", 1:8)
group   <- rep(c("control", "treated"), each = 4)
subject <- rep(c("P1", "P2", "P3", "P4"), times = 2)   # each subject: one of each group

## baseline mean expression per gene, with a low-expression tail
base_mu <- exp(rnorm(n_genes, mean = 7.5, sd = 1.5))
base_mu[(n_genes - 39):n_genes] <- runif(40, 0, 14)

## true log2 fold changes: first 60 genes differentially expressed, half up half down
lfc <- numeric(n_genes)
lfc[seq_len(n_true_de)] <- rep(c(1, -1), length.out = n_true_de) * runif(n_true_de, 1.2, 3)

## per-subject offset, so the paired/blocked analyses have a real effect to remove
subj_off <- stats::setNames(runif(4, 0.85, 1.18), c("P1", "P2", "P3", "P4"))

counts_mat <- sapply(seq_along(samples), function(j) {
  mu <- base_mu * subj_off[subject[j]]
  if (group[j] == "treated") mu <- mu * 2^lfc
  rnbinom(n_genes, mu = mu, size = 1 / 0.15)     # negative binomial, dispersion 0.15
})
dimnames(counts_mat) <- list(sprintf("ENSG%011d", seq_len(n_genes)), samples)

## the htseq-count summary rows that filtering is expected to strip
special <- c("__no_feature", "__ambiguous", "__too_low_aQual",
             "__not_aligned", "__alignment_not_unique")
counts_mat <- rbind(counts_mat,
                    matrix(sample(20000:90000, length(special) * 8, replace = TRUE),
                           nrow = length(special), dimnames = list(special, samples)))

samp_info_raw <- data.frame(SampleName = samples, Group = group,
                            Subject = subject, stringsAsFactors = FALSE)

dim(counts_mat)
head(counts_mat[, 1:4])
samp_info_raw
```

::: {style="text-align: justify;"}
Real analyses usually begin from a counts file rather than an in-memory matrix. To demonstrate that entry point we write the matrix to a temporary file and read it back with `diff_expr_read_counts()`, which exercises exactly the same code path. `diff_expr_get_samp_info()` first standardises the sample sheet to the conventions the package expects:
:::

```{r}
#| label: read-from-file
counts_file <- file.path(tempdir(), "example_counts.tsv")
write.table(counts_mat, counts_file, sep = "\t", quote = FALSE, col.names = NA)

samp.info <- diff_expr_get_samp_info(samp_info_raw,
                                     samples = "SampleName",
                                     groups  = "Group")
counts <- diff_expr_read_counts(counts_file, samp.info)
dim(counts)          # the htseq-count summary rows are still present here
head(counts[, 1:4])
```

## Choosing an analysis mode {#sec-modes}

::: {style="text-align: justify;"}
Three arguments together determine how an analysis is carried out: `pairs`, `block` and `do.voom`. Rather than acting independently, they resolve into one of a small number of well-defined modes. The distinction that matters most is **how the `pairs` column enters the model**:

-   **not used at all** --- a simple means model `~0 + groups`, with contrasts formed explicitly by `makeContrasts()`.
-   **as a fixed effect** (`pairs` given, `block = FALSE`) --- an additive model `~pairs + groups` with an intercept. Group coefficients are then already expressed relative to the control group and are extracted directly by coefficient name.
-   **as a correlation block** (`block = TRUE`) --- the means model again, but with the subject effect modelled through `limma::duplicateCorrelation()`.

The second axis is simply which engine performs the test: **edgeR**'s generalised linear models, or **limma** after a `voom` transformation, selected with `do.voom`.

These two axes give five valid combinations. The only constraint is that a blocked design forces the **limma** engine, because **edgeR** has no equivalent of `duplicateCorrelation()`; if `block = TRUE` is combined with `do.voom = FALSE`, `voom` is enabled automatically and the reason is written to the log.
:::

| `pairs` | `block` | `do.voom` | Design | Engine | Role of `pairs` |
|---------|---------|-----------|--------|--------|-----------------|
| --- | `FALSE` | `FALSE` | `~0 + groups` | edgeR GLM | --- |
| --- | `FALSE` | `TRUE` | `~0 + groups` | limma/voom | --- |
| given | `FALSE` | `FALSE` | `~pairs + groups` | edgeR GLM | fixed effect |
| given | `FALSE` | `TRUE` | `~pairs + groups` | limma/voom | fixed effect |
| given | `TRUE` | forced `TRUE` | `~0 + groups` | limma/voom | correlation block |

## Running the whole pipeline

::: {style="text-align: justify;"}
In the simplest case a single call is enough. Note that `out.dir` is **required**: the package deliberately has no default output location, so that nothing is ever written to the working directory unintentionally. Here the results go to a temporary directory, which is also what you should do in your own examples and tests.
:::

```{r}
#| label: full-run
#| eval: false
out.dir <- file.path(tempdir(), "diffwrap_example")
dir.create(out.dir, showWarnings = FALSE)

res <- diffExpr(expr.dat  = counts_file,
                samp.info = samp_info_raw,
                samples   = "SampleName",
                groups    = "Group",
                control   = "control",
                analysis.name = "demo",
                out.dir   = out.dir,
                enr.do = FALSE)
```

::: {style="text-align: justify;"}
The chunk above is not evaluated when the vignette is built, simply because a full run writes a fair number of files and takes noticeably longer than is worthwhile during a package build; it needs no network access with these settings. Run it interactively to see the complete output. The manual walk-through in the next section *is* evaluated, and shows real differential expression results.

`diffExpr()` returns a list of the objects created along the way --- the `DGEList` or `voom` object, the model fits, and one annotated result table per contrast --- but it is first and foremost called for its side effects. Below `out.dir` you will find, per contrast, a full and a filtered differential expression table, a PDF of the M-A, volcano and p-value plots, heatmaps, and a Venn diagram plus intersection tables across contrasts.
:::

### Paired and blocked analyses

```{r}
#| label: paired-run
#| eval: false
# 'Subject' pairs each control sample with a treated sample from the same subject
res_paired <- diffExpr(expr.dat  = counts_file,
                       samp.info = samp_info_raw,
                       samples   = "SampleName",
                       groups    = "Group",
                       pairs     = "Subject",     # fixed effect, intercept design
                       control   = "control",
                       analysis.name = "demo_paired",
                       out.dir   = out.dir,
                       enr.do = FALSE)

# the same column used as a correlation block instead (forces voom)
res_blocked <- diffExpr(expr.dat  = counts_file,
                        samp.info = samp_info_raw,
                        samples   = "SampleName",
                        groups    = "Group",
                        pairs     = "Subject",
                        block     = TRUE,         # duplicateCorrelation
                        control   = "control",
                        analysis.name = "demo_blocked",
                        out.dir   = out.dir,
                        enr.do = FALSE)
```

## Running the steps manually

::: {style="text-align: justify;"}
The wrapper is convenient, but each step is exported and can be called on its own. This is useful when you want to inspect intermediate objects, or plug a step into a different workflow.
:::

```{r}
#| label: manual-filter
counts_f <- diff_expr_filter_counts(counts, samp.info, strict = TRUE)
dim(counts_f)   # low-expression genes and the __-rows have gone
```

```{r}
#| label: manual-model
groups <- stats::relevel(samp.info$Groups, ref = "control")

d <- edgeR::DGEList(counts = counts_f, group = groups)
d <- edgeR::calcNormFactors(d)

design    <- diff_expr_make_design(samp.info = samp.info, groups = groups)
contrasts <- diff_expr_make_contrasts(design = design, groups = groups)

design
contrasts
```

```{r}
#| label: manual-fit
fit.l <- diff_expr_fit(counts = counts_f, d = d, design = design,
                       do.voom = FALSE, quasi.likelihood = TRUE)

de <- edgeR::glmQLFTest(fit.l$fit, contrast = contrasts[, 1])
head(edgeR::topTags(de, n = 10))
```

::: {style="text-align: justify;"}
Because the ground truth is known, the result can be checked directly: the truly differentially expressed genes are the first sixty, `ENSG00000000001` to `ENSG00000000060`.
:::

```{r}
#| label: check-truth
tt      <- as.data.frame(edgeR::topTags(de, n = Inf))
top50   <- rownames(tt)[seq_len(50)]
n_true  <- sum(as.integer(sub("^ENSG0*", "", top50)) <= 60)
cat("Of the 50 top-ranked genes,", n_true, "are truly differentially expressed\n")
```

## Output, logging and verbosity

::: {style="text-align: justify;"}
A pipeline of this size produces a lot of progress information. Rather than writing it all to the console, `diffExpr()` always writes a complete log file to `out.dir` and prints only the major steps to the console. The console channel uses `message()`, so it can be silenced in the usual R ways.

-   `verbose = TRUE` (default) --- major workflow steps on the console, full detail in the log file.
-   `verbose = FALSE` --- nothing on the console; the log file is still written in full.
-   `verbose = "all"` --- mirror the complete log to the console as well, which is handy while debugging.

`suppressMessages()` works too. The log file is named after the analysis and is written next to the results; pass `log.file` to put it somewhere else.
:::

```{r}
#| label: logging
#| eval: false
res <- diffExpr(..., out.dir = out.dir, verbose = FALSE)          # quiet console
res <- diffExpr(..., out.dir = out.dir, verbose = "all")          # everything echoed
res <- diffExpr(..., out.dir = out.dir,
                log.file = file.path(out.dir, "my_run.log"))      # custom log location
```

## Annotation and enrichment {#sec-enrichment}

::: {style="text-align: justify;"}
Setting `biom.use = TRUE` annotates the result tables with information retrieved from Ensembl, and `enr.do = TRUE` runs over-representation and gene set enrichment analyses through **clusterProfiler**, **gprofiler2** and **topGO**.

Both require network access, and enrichment additionally requires the relevant organism annotation package (`org.Hs.eg.db` for human, `org.Mm.eg.db` for mouse). KEGG pathway enrichment needs Entrez identifiers and therefore only works together with `biom.use = TRUE`; `diffExpr()` checks this at the start of a run and stops with an explanatory message rather than failing later.

Enrichment runs as a separate stage that consumes the differential-expression results: over-representation (ORA) on the significant genes and gene-set enrichment (GSEA) on the ranked gene list, across the methods selected in `enr.methods` (@fig-enrichment).
:::

![Enrichment sub-pipeline. The per-contrast DE results feed over-representation (ORA) and gene-set enrichment (GSEA), run through the methods chosen in enr.methods. KEGG needs Entrez IDs and therefore biomart annotation. Results are written as tables and optional network plots.](diffwrap-enrichment.svg){#fig-enrichment}

::: {style="text-align: justify;"}
Neither step will produce anything meaningful for the simulated example data, whose gene identifiers are well-formed but fictitious. The chunk below is therefore shown for reference only.
:::

```{r}
#| label: enrichment
#| eval: false
res <- diffExpr(expr.dat  = counts_file,
                samp.info = samp_info_raw,
                samples   = "SampleName",
                groups    = "Group",
                control   = "control",
                analysis.name = "demo_annotated",
                out.dir   = out.dir,
                biom.use  = TRUE,
                biom.data.set = "hsapiens_gene_ensembl",
                enr.do = TRUE,
                enr.methods = c("clusterProfilerGO", "gProfileR"))
```

## Session information

```{r}
#| label: session-info
sessionInfo()
```
